---
title: "06e: P-Spline GAM Resistance via PyTorch with Bayesian UQ"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{06e: P-Spline GAM Resistance via PyTorch with Bayesian UQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

This vignette demonstrates the P-spline GAM resistance pipeline: a
differentiable B-spline model with penalized complexity learns nonlinear
covariate effects on log-resistance, optimized via Adam through the global
absorption circuit solver. Unlike the black-box MLP of vignette 06d, the
spline model produces interpretable partial-effect curves with smoothness
controlled by learnable penalty parameters.

Optionally, Bayesian uncertainty quantification (UQ) can be computed after MAP
estimation using ADVI, Langevin MCMC, or NUTS.

Like vignette 06d, this pipeline requires a 5-layer basis (4 NLCD + elevation).

## Setup

```{r paths}
library(DiffiScape)
library(terra)

obs_data_path  <- "path/to/train_points.rds"
nlcd_dir       <- "path/to/nlcd_data/"
elevation_path <- "path/to/elevation.tif"
output_dir     <- "output/spline_gam"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load data and build 5-layer basis

```{r load-data}
obs_points <- readRDS(obs_data_path)

nlcd_paths  <- list.files(nlcd_dir, pattern = "\\.tif$", full.names = TRUE)
basis_stack <- ds_create_basis(c(nlcd_paths, elevation_path), rescale = TRUE)
```

## Initialise PyTorch

```{r init-torch}
ds_torch_setup()
```

## Verify spline gradient (optional)

```{r verify-gradient}
verify_spline_gradient(
  basis_stack           = basis_stack,
  solver                = "global_absorption",
  absorption            = 0.0005,
  n_knots               = 3L,
  spline_degree         = 3L,
  include_interactions  = FALSE
)
```

## Configure

Key choices:

- `model_type = "spline_gam"` — P-spline resistance instead of MLP.
- `n_knots = 3L` — cubic B-spline with 3 internal knots per covariate (5 basis
  functions per covariate after accounting for boundary knots).
- `penalty_scale = 3.0` — global smoothing multiplier; higher = smoother.
- `absorption = 0.0005` — lower absorption makes the circuit more global.
- `include_interactions = FALSE` — marginal smooths only; set `TRUE` for
  tensor-product interaction terms (much slower).

```{r config}
cfg <- default_torch_config()

# Solver
cfg$solver                   <- "global_absorption"
cfg$absorption               <- 0.0005
cfg$source_spacing           <- 1L
cfg$source_from_resistance   <- TRUE

# Spline architecture
cfg$model_type               <- "spline_gam"
cfg$n_knots                  <- 3L
cfg$spline_degree            <- 3L
cfg$include_interactions     <- FALSE
cfg$penalty_scale            <- 3.0
cfg$lambda_init_marginal     <- 2.5   # log initial smoothing param per covariate
cfg$lambda_init_interaction  <- 2.5
cfg$lambda_min               <- 0.1   # floor: λ = lambda_min + exp(log_λ)

# Intensity model (parametric γ·log(1+C); set intensity_spline=TRUE for spline)
cfg$intensity_spline         <- FALSE

# Optimiser
cfg$lr                       <- 0.07
cfg$weight_decay             <- 0.0
cfg$n_epochs                 <- 500L
cfg$patience                 <- 100L
cfg$reg_mean                 <- 0.5
cfg$reg_var                  <- 0.0
cfg$log_R_baseline           <- 3.0

cfg$seed                     <- 42L
cfg$device                   <- "auto"
```

## MAP estimation

```{r run-map}
results <- run_torch_pipeline(
  basis_stack = basis_stack,
  obs_points  = obs_points,
  config      = cfg,
  output_dir  = output_dir
)

cat("Best log-likelihood:", results$best_loglik, "\n")
cat("Best epoch:", results$best_epoch, "/", results$n_epochs_run, "\n")
```

## Inspect partial effects

`results$partial_effects` is a named list (one entry per basis layer). Each
entry contains the 1-D covariate grid and the estimated log-resistance effect
with 50 % and 95 % pointwise intervals (populated after UQ, or NA before UQ).

```{r partial-effects}
for (nm in names(results$partial_effects)) {
  pe <- results$partial_effects[[nm]]
  plot(pe$grid, pe$effect, type = "l", main = nm,
       xlab = nm, ylab = "log-resistance partial effect")
}
```

## Save MAP outputs

```{r save-map}
terra::writeRaster(results$resistance_raster,
                   file.path(output_dir, "optimized_resistance.tif"), overwrite = TRUE)
terra::writeRaster(results$connectivity_raster,
                   file.path(output_dir, "optimized_connectivity.tif"), overwrite = TRUE)
terra::writeRaster(results$intensity_raster,
                   file.path(output_dir, "predicted_intensity.tif"), overwrite = TRUE)
```

## Bayesian UQ — ADVI (fastest)

ADVI fits a mean-field Gaussian variational approximation to the posterior.
It is faster than MCMC but provides an approximate posterior.

```{r advi}
advi_results <- run_advi(
  basis_stack     = basis_stack,
  obs_points      = obs_points,
  model_dir       = output_dir,
  n_samples       = 1000L,
  max_iter        = 2000L,
  lr              = 0.01,
  n_elbo_samples  = 8L,
  full_rank       = FALSE,       # mean-field; TRUE = full-rank covariance
  patience        = 100L,
  fix_smoothing   = TRUE,        # hold smoothing params fixed at MAP
  fix_intensity   = FALSE
)
```

## Bayesian UQ — Langevin MCMC (moderate cost)

Preconditioned Metropolis-adjusted Langevin algorithm (MALA). Targets ~80 %
acceptance. `fix_smoothing = TRUE` holds the spline penalty parameters at
their MAP estimates, reducing the parameter space.

```{r langevin}
mcmc_results <- run_bayesian_sampling(
  basis_stack     = basis_stack,
  obs_points      = obs_points,
  model_dir       = output_dir,
  n_samples       = 1000L,
  burn_in         = 500L,
  thin            = 10L,
  step_size       = 2e-6,
  precondition    = TRUE,
  fix_smoothing   = TRUE,
  fix_intensity   = FALSE,
  use_mala        = TRUE,
  target_accept   = 0.80
)
```

## Bayesian UQ — NUTS / HMC (highest quality)

No-U-Turn Sampler with adaptive mass matrix. Most expensive, but produces
the most accurate posterior.

```{r nuts}
hmc_results <- run_bayesian_sampling_hmc(
  basis_stack       = basis_stack,
  obs_points        = obs_points,
  model_dir         = output_dir,
  n_samples         = 500L,
  warmup            = 1000L,
  max_treedepth     = 7L,
  target_accept     = 0.80,
  adapt_mass_matrix = TRUE,
  fix_smoothing     = TRUE,
  fix_intensity     = FALSE
)
```

## Save full summary

```{r save}
saveRDS(
  list(
    effective_loglinear = results$effective_loglinear,
    alpha               = results$alpha,
    gamma               = results$gamma,
    partial_effects     = results$partial_effects,
    best_loglik         = results$best_loglik,
    total_time          = results$total_time,
    uq_advi             = advi_results,
    uq_mcmc             = mcmc_results,
    uq_hmc              = hmc_results
  ),
  file.path(output_dir, "final_summary.rds")
)
```
