---
title: "Profile-Likelihood GAM via Gradient Optimization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profile-Likelihood GAM via Gradient Optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

This vignette shows the profile-likelihood GAM pipeline: gradient-based
optimization of the resistance parameters where each objective evaluation
refits a full spatial GAM intensity model (via `mgcv::bam`). The GAM captures
nonlinear spatial variation in intensity beyond what the connectivity surface
explains alone.

The gradient of the profile log-likelihood is obtained by the envelope theorem:
only the partial derivative with respect to connectivity carries a non-zero
contribution at the GAM optimum, so the chain rule through the circuit solver
gives exact gradients without differentiating through the GAM solver.

**Note:** Use `icfg$include_spatial_re <- FALSE` and `icfg$bam_discrete <- FALSE`
during optimization (required for the envelope theorem to hold); you can refit
with spatial RE and `discrete = TRUE` for the final intensity estimate.

## Setup

```{r paths}
library(DiffiScape)
library(terra)

obs_data_path <- "path/to/train_points.rds"
nlcd_dir      <- "path/to/nlcd_data/"
output_dir    <- "output/gam_profile"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load data

```{r load-data}
obs_points  <- readRDS(obs_data_path)
basis_stack <- ds_create_basis(nlcd_dir, pattern = "*.tif", rescale = TRUE)
```

## Initialise JAX

```{r init-jax}
ds_jax_setup()
```

## Configure

The key intensity config settings for the profile-likelihood approach:

- `covariate_type = "smooth"` — spatial tensor-product smooth via `bam()`
- `include_spatial_re = FALSE` — disable spatial RE during optimization (avoid
  saturation; re-enable for final fit)
- `k_connectivity = 10L` — knots for connectivity smooth term

```{r config}
cfg <- default_optimizer_config()
cfg$distribution  <- "gam"
cfg$connectivity  <- list(radius = 13L, block_size = 5L)

icfg <- default_intensity_config()
icfg$covariate_type      <- "smooth"
icfg$k_connectivity      <- 10L
icfg$include_spatial_re  <- FALSE    # required for envelope theorem correctness
```

## Optimise (profile-likelihood outer loop, GAM inner loop)

```{r optimize}
opt <- ds_optimize(
  basis_stack      = basis_stack,
  obs_points       = obs_points,
  config           = cfg,
  intensity_config = icfg,
  output_dir       = output_dir,
  solver           = "gradient"   # gradient-based via JAX
)

cat("Best profile log-likelihood:", opt$best_loglik, "\n")
cat("Best resistance params:\n")
print(opt$best_params)
```

## Final connectivity

```{r connectivity}
resistance   <- create_resistance_surface(opt$best_params, basis_stack)
jax_result   <- ds_jax_connectivity(resistance, radius = 13L, block_size = 5L)
connectivity <- jax_result$cum_current

terra::writeRaster(resistance,   file.path(output_dir, "optimized_resistance.tif"),  overwrite = TRUE)
terra::writeRaster(connectivity, file.path(output_dir, "optimized_connectivity.tif"), overwrite = TRUE)
```

## Final GAM intensity fit

For the final intensity fit, re-enable the spatial random effect and use a
fuller GAM specification. This fit is for inference and prediction, not for
gradient computation, so the envelope-theorem constraints no longer apply.

```{r fit-intensity}
icfg_final <- icfg
icfg_final$include_spatial_re <- TRUE   # enable spatial smooth for final fit

intensity_fit <- ds_fit_intensity(
  opt_result         = opt,
  basis_stack        = basis_stack,
  obs_points         = obs_points,
  connectivity_settings = list(radius = 13L, block_size = 5L),
  intensity_config   = icfg_final,
  solver             = "gradient"
)

cat("Final GAM log-likelihood:", intensity_fit$loglik, "\n")
```

## Predict and diagnose

```{r predict}
intensity_map <- ds_predict(intensity_fit, connectivity, intensity_config = icfg_final)
terra::writeRaster(intensity_map, file.path(output_dir, "predicted_intensity.tif"), overwrite = TRUE)
```

```{r diagnose}
diag <- ds_diagnose(
  intensity_fit    = intensity_fit,
  obs_points       = obs_points,
  connectivity     = connectivity,
  intensity_config = icfg_final,
  plot             = TRUE
)
```

## Save

```{r save}
saveRDS(
  list(
    resistance_params    = opt$best_params,
    intensity_params     = intensity_fit$intensity_params,
    loglik               = intensity_fit$loglik,
    convergence          = opt$convergence
  ),
  file.path(output_dir, "final_summary.rds")
)
```
