---
title: "Gradient-Based Optimization via JAX"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gradient-Based Optimization via JAX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

This vignette demonstrates the gradient-based pipeline: L-BFGS optimization of
resistance parameters using exact gradients computed via JAX (JAX automatic
differentiation) through a differentiable circuit solver. Compared to the
surrogate approach in vignette 06, this pipeline converges in minutes rather
than hours, but is sensitive to initialisation and local optima.

Internally DiffiScape uses a logit reparameterization of the bounded resistance
parameters and an adaptive standardization restart loop to improve convergence.

## Setup

```{r paths}
library(DiffiScape)
library(terra)

obs_data_path <- "path/to/train_points.rds"
nlcd_dir      <- "path/to/nlcd_data/"
output_dir    <- "output/gradient"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load data

```{r load-data}
obs_points  <- readRDS(obs_data_path)
basis_stack <- ds_create_basis(nlcd_dir, pattern = "*.tif", rescale = TRUE)
```

## Initialise JAX

```{r init-jax}
ds_jax_setup()
```

## Configure

The negative binomial intensity model is more robust than Poisson when counts
are overdispersed. Connectivity is computed by the JAXScape circuit-theory
solver with JAX automatic differentiation gradients.

```{r config}
cfg <- default_optimizer_config()
cfg$distribution  <- "negbin"
cfg$connectivity  <- list(radius = 15L, block_size = 5L)

icfg <- default_intensity_config()
```

## Warm-start from surrogate results (optional)

If you ran the surrogate pipeline (vignette 06), you can warm-start gradient
optimization from those results to save time.

```{r warm-start}
surrogate_summary <- readRDS("output/surrogate/final_summary.rds")
theta_init <- surrogate_summary$resistance_params   # named numeric vector
```

## Optimise

`solver = "gradient"` selects the JAX gradient backend. The optimiser
applies L-BFGS with logit reparameterization and restarts if connectivity
standardization constants shift during the run.

```{r optimize}
opt <- ds_optimize(
  basis_stack      = basis_stack,
  obs_points       = obs_points,
  config           = cfg,
  intensity_config = icfg,
  output_dir       = output_dir,
  solver           = "gradient"
)

cat("Best log-likelihood:", opt$best_loglik, "\n")
cat("Convergence message:", opt$convergence, "\n")
print(opt$best_params)
```

## Final connectivity and intensity

```{r connectivity}
resistance   <- create_resistance_surface(opt$best_params, basis_stack)
jax_result   <- ds_jax_connectivity(resistance, radius = 15L, block_size = 5L)
connectivity <- jax_result$cum_current

terra::writeRaster(resistance,   file.path(output_dir, "optimized_resistance.tif"),  overwrite = TRUE)
terra::writeRaster(connectivity, file.path(output_dir, "optimized_connectivity.tif"), overwrite = TRUE)
```

```{r fit-intensity}
intensity_fit <- ds_fit_intensity(
  opt_result         = opt,
  basis_stack        = basis_stack,
  obs_points         = obs_points,
  omniscape_settings = list(radius = 15L, block_size = 5L),
  intensity_config      = icfg,
  solver                = "gradient"
)

cat("Intensity log-likelihood:", intensity_fit$loglik, "\n")
```

```{r predict}
intensity_map <- ds_predict(intensity_fit, connectivity)

terra::writeRaster(intensity_map, file.path(output_dir, "predicted_intensity.tif"), overwrite = TRUE)
```

## Diagnostics

```{r diagnose}
diag <- ds_diagnose(intensity_fit, obs_points, connectivity, plot = TRUE)
```

## Extract connectivity at observations

```{r extract-connectivity}
C_obs <- extract_connectivity(connectivity, obs_points)
cat("Mean connectivity at observations:", mean(C_obs, na.rm = TRUE), "\n")
```

## Save

```{r save}
saveRDS(
  list(
    resistance_params = opt$best_params,
    intensity_params  = intensity_fit$intensity_params,
    loglik            = intensity_fit$loglik,
    convergence       = opt$convergence,
    n_restarts        = opt$n_restarts
  ),
  file.path(output_dir, "final_summary.rds")
)
```
