---
title: "Inverse Reinforcement Learning Resistance via PyTorch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inverse Reinforcement Learning Resistance via PyTorch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

This vignette fits an inverse-reinforcement-learning (IRL) resistance model to
a synthetic 100 x 100 urban landscape. Its differentiable chain is:

```
covariates -> learned reward -> soft value iteration -> long-range value
           -> resistance -> differentiable circuit connectivity -> PPP intensity
```

High value produces low resistance. Circuit theory remains the forward model:
IRL learns a reward and propagates its value before the circuit solve, rather
than replacing the circuit model. This is useful when occurrence points suffice
but movement trajectories are unavailable.

## Simulate the urban landscape and point process

The landscape matches the MLP and P-spline examples: four park blobs, a
sinusoidal road grid, canopy cover, impervious surface, and a twice-smoothed
random spatial effect. The direct log-linear truth keeps the example
comparable with those vignettes; IRL fits it through propagated value rather
than interpreting the two covariates as local fitted coefficients.

```{r data}
library(DiffiScape)
library(terra)

withr::local_seed(42)
template <- terra::rast(
  nrows = 100L, ncols = 100L,
  xmin = 0, xmax = 10000, ymin = 0, ymax = 10000, crs = "EPSG:3857"
)
coords <- terra::crds(template, df = TRUE)
x <- (coords$x - min(coords$x)) / diff(range(coords$x))
y <- (coords$y - min(coords$y)) / diff(range(coords$y))
layer <- function(values) {
  raster <- template
  terra::values(raster) <- values
  raster
}

# Four irregular parks embedded in an urban matrix, plus a simple road grid.
parks <- pmin(1, exp(-((x - 0.20)^2 + (y - 0.76)^2) / 0.012) +
                 exp(-((x - 0.72)^2 + (y - 0.72)^2) / 0.020) +
                 exp(-((x - 0.30)^2 + (y - 0.25)^2) / 0.010) +
                 exp(-((x - 0.78)^2 + (y - 0.30)^2) / 0.014))
roads <- as.numeric(abs(sin(8 * pi * x)) < 0.08 |
                    abs(sin(8 * pi * y)) < 0.08)
canopy <- pmin(1, pmax(0, 0.08 + 0.90 * parks - 0.20 * roads +
                          rnorm(terra::ncell(template), sd = 0.04)))
impervious <- pmin(1, pmax(0, 0.15 + 0.25 * (1 - parks) + 0.40 * roads + 0.25 * x +
                              rnorm(terra::ncell(template), sd = 0.04)))
basis_stack <- create_basis_stack(list(
  canopy = layer(canopy),
  impervious = layer(impervious)
), rescale = TRUE)

# Smooth white noise twice to create an omitted spatial log-intensity effect.
spatial_effect <- terra::focal(layer(rnorm(terra::ncell(template))),
                               w = matrix(1, 9, 9), fun = "mean",
                               na.rm = TRUE, fillvalue = 0)
spatial_effect <- terra::focal(spatial_effect, w = matrix(1, 9, 9),
                               fun = "mean", na.rm = TRUE, fillvalue = 0)
spatial_effect <- 0.35 * as.numeric(scale(terra::values(spatial_effect)))

true_log_resistance <- -1 * canopy + 1 * impervious
true_resistance <- layer(exp(true_log_resistance))
ds_torch_setup()
circuit_solver <- DiffiScape:::.ds_env$torch_module$`_get_circuit_module`()
truth <- circuit_solver$solve_circuit_absorption(
  terra::as.matrix(true_resistance, wide = TRUE), absorption = 0.20,
  source_spacing = 1L, source_from_resistance = TRUE
)
true_connectivity <- layer(as.numeric(reticulate::py_to_r(truth$current_density)))
log_lambda <- 2 * log1p(terra::values(true_connectivity)) + spatial_effect
sampling_weight <- exp(log_lambda - max(log_lambda))
obs_cells <- sample(seq_len(terra::ncell(template)), 1000L,
                    replace = TRUE, prob = sampling_weight)
obs_points <- coords[obs_cells, c("x", "y")]
output_dir <- file.path(tempdir(), "diffiscape-irl")
```

## Check the IRL gradient and fit

The gradient check exercises the reward-to-value-to-circuit chain on a small
crop. `beta` is the soft value-iteration temperature, `gamma_d` discounts (or
leaks) distant value, and `n_value_iter` sets the planning-horizon
approximation. More value-iteration steps extend that approximation but also
increase autograd memory.

```{r fit}
gradient_check <- verify_irl_gradient(
  basis_stack, crop_size = 12L, solver = "global_absorption",
  absorption = 0.20, hidden_dim = 4L, n_hidden_layers = 1L,
  beta = 1.0, gamma_d = 0.9, n_value_iter = 20L,
  value_scale_init = 1.0
)
stopifnot(isTRUE(gradient_check$pass))

cfg <- default_torch_config()
cfg$solver <- "global_absorption"
cfg$absorption <- 0.20
cfg$source_spacing <- 1L
cfg$model_type <- "irl"
cfg$hidden_dim <- 4L
cfg$n_hidden_layers <- 1L
cfg$beta <- 1.0
cfg$gamma_d <- 0.9
cfg$n_value_iter <- 20L
cfg$value_scale_init <- 1.0
cfg$lr <- 0.005
cfg$n_epochs <- 10L
cfg$patience <- 5L
cfg$warmup_epochs <- 1L
cfg$reg_mean <- 0.2
cfg$reg_var <- 0.0
cfg$reg_skip <- 0.05
cfg$log_R_baseline <- 0.0
cfg$seed <- 42L
cfg$device <- "cpu"

results <- run_torch_pipeline(
  basis_stack = basis_stack, obs_points = obs_points,
  config = cfg, output_dir = output_dir,
  warm_start_theta = c(0, -0.5, 0.5)
)
```

## Diagnostics, maps, and the public wrapper

The fitted model deliberately omits the simulated spatial effect, so its maps
should recover broad resistance and connectivity structure rather than every
local intensity fluctuation.

```{r diagnostics, fig.width=12, fig.height=4, dpi=96}
stopifnot(
  all(is.finite(results$loss_history)),
  all(is.finite(terra::values(results$resistance_raster))),
  all(is.finite(terra::values(results$connectivity_raster))),
  all(is.finite(terra::values(results$intensity_raster))),
  all(results$loss_history > 0),
  all(terra::values(results$resistance_raster) > 0),
  all(terra::values(results$connectivity_raster) > 0),
  all(terra::values(results$intensity_raster) > 0)
)

plot(results$loss_history, type = "l", xlab = "Epoch", ylab = "Objective",
     main = "Training objective")
op <- par(mfrow = c(1, 3))
plot(true_resistance, main = "True resistance")
plot(results$resistance_raster, main = "Fitted resistance")
plot(results$connectivity_raster, main = "Fitted connectivity")
par(op)

model <- irl_resistance_model(results, basis_stack)
wrapped_resistance <- predict(model)
stopifnot(isTRUE(all.equal(
  terra::values(wrapped_resistance),
  terra::values(results$resistance_raster)
)))
```

## Save compact results

```{r save}
saveRDS(
  list(
    true_log_resistance = true_log_resistance,
    true_connectivity = true_connectivity,
    best_loglik = results$best_loglik,
    best_epoch = results$best_epoch,
    total_time = results$total_time,
    beta = cfg$beta,
    gamma_d = cfg$gamma_d,
    n_value_iter = cfg$n_value_iter
  ),
  file.path(output_dir, "torch_irl_optimization_results.rds")
)
```
