---
title: "P-Spline GAM Resistance via PyTorch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{P-Spline GAM Resistance via PyTorch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
```

This vignette uses the same synthetic urban point process as the neural MLP
vignette, but fits an interpretable P-spline resistance model. It has two
input rasters: canopy cover for interspersed urban parks and impervious
surface for roads and built areas.

## Simulate a 100 x 100 urban landscape and 1,000 points

The true resistance model is \(\log R = -1 * \mathrm{canopy} + 1 *
\mathrm{impervious}\). We calculate its circuit connectivity, then simulate
from \(\log \lambda = 2\log(1 + C) + u(s)\), with \(u(s)\) a smooth spatially
autocorrelated random effect. Sampling 1,000 cells conditional on the total
count is the conditional Poisson point-process construction.

```{r data}
library(DiffiScape)
library(terra)

withr::local_seed(42)
template <- terra::rast(
  nrows = 100L, ncols = 100L,
  xmin = 0, xmax = 10000, ymin = 0, ymax = 10000, crs = "EPSG:3857"
)
coords <- terra::crds(template, df = TRUE)
x <- (coords$x - min(coords$x)) / diff(range(coords$x))
y <- (coords$y - min(coords$y)) / diff(range(coords$y))
layer <- function(values) {
  raster <- template
  terra::values(raster) <- values
  raster
}

parks <- pmin(1, exp(-((x - 0.20)^2 + (y - 0.76)^2) / 0.012) +
                 exp(-((x - 0.72)^2 + (y - 0.72)^2) / 0.020) +
                 exp(-((x - 0.30)^2 + (y - 0.25)^2) / 0.010) +
                 exp(-((x - 0.78)^2 + (y - 0.30)^2) / 0.014))
roads <- as.numeric(abs(sin(8 * pi * x)) < 0.08 |
                    abs(sin(8 * pi * y)) < 0.08)
canopy <- pmin(1, pmax(0, 0.08 + 0.90 * parks - 0.20 * roads +
                          rnorm(terra::ncell(template), sd = 0.04)))
impervious <- pmin(1, pmax(0, 0.15 + 0.25 * (1 - parks) + 0.40 * roads + 0.25 * x +
                              rnorm(terra::ncell(template), sd = 0.04)))
basis_stack <- create_basis_stack(list(
  canopy = layer(canopy),
  impervious = layer(impervious)
), rescale = TRUE)

spatial_effect <- terra::focal(layer(rnorm(terra::ncell(template))),
                               w = matrix(1, 9, 9), fun = "mean",
                               na.rm = TRUE, fillvalue = 0)
spatial_effect <- terra::focal(spatial_effect, w = matrix(1, 9, 9),
                               fun = "mean", na.rm = TRUE, fillvalue = 0)
spatial_effect <- 0.35 * as.numeric(scale(terra::values(spatial_effect)))

true_log_resistance <- -1 * canopy + 1 * impervious
true_resistance <- layer(exp(true_log_resistance))
ds_torch_setup()
circuit_solver <- DiffiScape:::.ds_env$torch_module$`_get_circuit_module`()
truth <- circuit_solver$solve_circuit_absorption(
  terra::as.matrix(true_resistance, wide = TRUE), absorption = 0.20,
  source_spacing = 1L, source_from_resistance = TRUE
)
true_connectivity <- layer(as.numeric(reticulate::py_to_r(truth$current_density)))
log_lambda <- 2 * log1p(terra::values(true_connectivity)) + spatial_effect
sampling_weight <- exp(log_lambda - max(log_lambda))
obs_cells <- sample(seq_len(terra::ncell(template)), 1000L,
                    replace = TRUE, prob = sampling_weight)
obs_points <- coords[obs_cells, c("x", "y")]
output_dir <- file.path(tempdir(), "diffiscape-spline-gam")
```

## Fit a small P-spline model

Only marginal smooths are included. The gradient check uses a 12 x 12 crop;
the map fit uses all 10,000 cells but limits optimisation to 10 epochs and
uses a relatively local absorption term so the vignette remains viable on a
CPU-only machine. It starts from `c(0, -0.5, 0.5)`, a conservative
sign-correct approximation rather than the simulated true coefficients, to
keep the short optimisation scientifically interpretable.

```{r fit}
gradient_check <- verify_spline_gradient(
  basis_stack, crop_size = 12L, solver = "global_absorption",
  absorption = 0.20, n_knots = 3L, spline_degree = 3L,
  include_interactions = FALSE
)
stopifnot(isTRUE(gradient_check$pass))

cfg <- default_torch_config()
cfg$solver <- "global_absorption"
cfg$absorption <- 0.20
cfg$source_spacing <- 1L
cfg$model_type <- "spline_gam"
cfg$n_knots <- 3L
cfg$spline_degree <- 3L
cfg$include_interactions <- FALSE
cfg$penalty_scale <- 3.0
cfg$lambda_init_marginal <- 2.5
cfg$lambda_min <- 0.1
cfg$lr <- 0.005
cfg$n_epochs <- 10L
cfg$patience <- 5L
cfg$warmup_epochs <- 1L
cfg$reg_mean <- 0.2
cfg$reg_var <- 0.0
cfg$log_R_baseline <- 0.0
cfg$seed <- 42L
cfg$device <- "cpu"

results <- run_torch_pipeline(
  basis_stack = basis_stack, obs_points = obs_points,
  config = cfg, output_dir = output_dir,
  warm_start_theta = c(0, -0.5, 0.5)
)
```

## Diagnostic, map, and partial-effect checks

The fitted model deliberately lacks the simulated spatial random effect. Its
smooth effects should nevertheless show lower fitted resistance with canopy
and higher fitted resistance with impervious cover, while the maps capture
the broad park-and-urban pattern.

```{r diagnostics, fig.width=12, fig.height=4, dpi=96}
stopifnot(
  all(is.finite(results$loss_history)),
  all(is.finite(terra::values(results$resistance_raster))),
  all(is.finite(terra::values(results$connectivity_raster))),
  all(is.finite(terra::values(results$intensity_raster))),
  all(terra::values(results$resistance_raster) > 0),
  all(terra::values(results$intensity_raster) > 0)
)
print(data.frame(
  coefficient = c("canopy", "impervious"),
  true_log_resistance_effect = c(-1, 1),
  fitted_effective_loglinear = results$effective_loglinear[-1]
))

plot(results$loss_history, type = "l", xlab = "Epoch", ylab = "Objective",
     main = "Training objective")
op <- par(mfrow = c(1, 3))
plot(true_resistance, main = "True resistance")
plot(results$resistance_raster, main = "Fitted resistance")
plot(results$connectivity_raster, main = "Fitted connectivity")
par(op)
```

```{r partial-effects, fig.width=8, fig.height=4, dpi=96}
op <- par(mfrow = c(1, 2))
for (index in names(results$partial_effects)) {
  partial_effect <- results$partial_effects[[index]]
  layer_name <- names(basis_stack)[as.integer(index) + 1L]
  plot(partial_effect$grid, partial_effect$effect, type = "l",
       xlab = layer_name, ylab = "Log-resistance contribution",
       main = layer_name)
}
par(op)
```

```{r save}
saveRDS(
  list(
    true_log_resistance = true_log_resistance,
    true_connectivity = true_connectivity,
    effective_loglinear = results$effective_loglinear,
    partial_effects = results$partial_effects,
    best_loglik = results$best_loglik,
    total_time = results$total_time
  ),
  file.path(output_dir, "torch_spline_optimization_results.rds")
)
```
