---
title: "Bayesian Surrogate Optimization with Omniscape"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Surrogate Optimization with Omniscape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

This vignette walks through the reference DiffiScape pipeline: Bayesian surrogate
(Gaussian Process) optimization of resistance parameters, with connectivity computed
by Omniscape.jl. The intensity surface is modelled with a spatial GAM (`mgcv::bam`),
and resistance uncertainty is quantified via a Laplace-hybrid posterior.

This is the most compute-intensive pipeline (~hours), but it requires no automatic
differentiation and produces well-calibrated uncertainty estimates.

## Setup

Set your data paths before running. The deer GPS data must be a data frame with
`x` and `y` columns in a projected CRS. NLCD rasters should be single-layer
`.tif` files in one directory (canopy, impervious surface, water, fence / open
land, etc.).

```{r paths}
library(DiffiScape)
library(terra)

obs_data_path <- "path/to/train_points.rds"   # data.frame with x, y
nlcd_dir      <- "path/to/nlcd_data/"
output_dir    <- "output/surrogate"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load data

```{r load-data}
# GPS observation points
obs_points <- readRDS(obs_data_path)          # data.frame(x, y)
# ds_load_data() accepts CSV / shapefile / GeoPackage as alternatives:
# obs_points <- ds_load_data("points.csv", x_col = "lon", y_col = "lat")

# Environmental basis functions (rescaled to [0, 1])
basis_stack <- ds_create_basis(nlcd_dir, pattern = "*.tif", rescale = TRUE)
cat("Basis layers:", names(basis_stack), "\n")
```

## Initialise Julia / Omniscape

```{r init-julia}
ds_init_julia()   # one-time setup per session
```

## Configure the optimiser

The outer loop is a Gaussian Process surrogate with Thompson Sampling acquisition.
The inner loop fits a spatial GAM intensity model (`distribution = "gam"`).

```{r config}
cfg <- default_optimizer_config()
cfg$n_init        <- 30L          # Latin Hypercube seed designs
cfg$n_iter        <- 70L          # GP surrogate iterations
cfg$seed          <- 42L
cfg$distribution  <- "gam"        # "negbin", "poisson", or "gam"
cfg$omniscape     <- list(radius = 25L, block_size = 5L, cleanup = TRUE)

icfg <- default_intensity_config()
icfg$covariate_type      <- "smooth"   # spatial GAM smooth terms
icfg$k_connectivity      <- 10L
icfg$include_spatial_re  <- TRUE
```

## Optimise resistance parameters

Returns the best resistance parameter vector `opt$best_params` (named numeric:
`r_0`, `z_1`, …, `z_K`) and the corresponding log-likelihood.

```{r optimize}
opt <- ds_optimize(
  basis_stack     = basis_stack,
  obs_points      = obs_points,
  config          = cfg,
  intensity_config = icfg,
  output_dir      = output_dir,
  solver          = "surrogate"
)

cat("Best log-likelihood:", opt$best_loglik, "\n")
cat("Best resistance params:\n")
print(opt$best_params)
```

## Compute final connectivity

```{r connectivity}
resistance  <- create_resistance_surface(opt$best_params, basis_stack)
omni        <- run_omniscape(resistance, radius = 25L, block_size = 5L, cleanup = TRUE)
connectivity <- omni$cum_current

terra::writeRaster(resistance,   file.path(output_dir, "optimized_resistance.tif"),  overwrite = TRUE)
terra::writeRaster(connectivity, file.path(output_dir, "optimized_connectivity.tif"), overwrite = TRUE)
```

## Fit intensity model

Refits the GAM at the optimised resistance parameters to obtain coefficient
estimates, standard errors, and final log-likelihood.

```{r fit-intensity}
intensity_fit <- ds_fit_intensity(
  opt_result       = opt,
  basis_stack      = basis_stack,
  obs_points       = obs_points,
  omniscape_settings = list(radius = 25L, block_size = 5L),
  intensity_config = icfg
)

cat("Intensity log-likelihood:", intensity_fit$loglik, "\n")
```

## Predict intensity surface

```{r predict}
intensity_map <- ds_predict(
  intensity_fit = intensity_fit,
  connectivity  = connectivity,
  intensity_config = icfg
)

terra::writeRaster(intensity_map, file.path(output_dir, "predicted_intensity.tif"), overwrite = TRUE)
```

## Posterior inference

A Laplace approximation followed by Monte Carlo composition yields a posterior
distribution over resistance parameters. `n_draws` is the number of posterior
samples; the original analysis used 5 000.

```{r posterior}
post <- ds_posterior(
  opt_result       = opt,
  basis_stack      = basis_stack,
  obs_points       = obs_points,
  n_draws          = 5000L,
  omniscape_settings = list(radius = 25L, block_size = 5L),
  intensity_config = icfg
)

print(post$summary)
write.csv(post$summary, file.path(output_dir, "posterior_summary.csv"), row.names = FALSE)
```

## Diagnostics

```{r diagnose}
diag <- ds_diagnose(
  intensity_fit = intensity_fit,
  obs_points    = obs_points,
  connectivity  = connectivity,
  intensity_config = icfg,
  plot          = TRUE
)

cat("Mean deviance residual:", diag$mean_deviance, "\n")
cat("Moran's I p-value:", diag$moran$p.value, "\n")
```

## Save summary

```{r save}
saveRDS(
  list(
    resistance_params = opt$best_params,
    intensity_params  = intensity_fit$intensity_params,
    loglik            = intensity_fit$loglik,
    posterior         = post$summary
  ),
  file.path(output_dir, "final_summary.rds")
)
```
