---
title: "Neural Network Resistance via PyTorch (MLP)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Neural Network Resistance via PyTorch (MLP)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
```

This vignette fits a small neural resistance model to a synthetic 100 x 100
urban landscape. Parks are represented by high canopy cover, while roads and
built areas have high impervious cover. The example is designed to exercise
the full differentiable circuit pipeline on a CPU, rather than to provide a
real ecological analysis.

## Simulate the urban landscape and point process

The two input rasters are canopy cover and impervious surface, each scaled to
[0, 1]. The data-generating resistance model is

\[\log R = -1 * \mathrm{canopy} + 1 * \mathrm{impervious}.\]

We solve the global-absorption circuit on that true resistance surface. The
point-process model is \(\log \lambda = 2\log(1 + C) + u(s)\), where \(u(s)\)
is a smooth, spatially autocorrelated random effect. Finally, we condition on
1,000 observed events and sample their cells in proportion to \(\lambda\).

```{r data}
library(DiffiScape)
library(terra)

withr::local_seed(42)
template <- terra::rast(
  nrows = 100L, ncols = 100L,
  xmin = 0, xmax = 10000, ymin = 0, ymax = 10000, crs = "EPSG:3857"
)
coords <- terra::crds(template, df = TRUE)
x <- (coords$x - min(coords$x)) / diff(range(coords$x))
y <- (coords$y - min(coords$y)) / diff(range(coords$y))
layer <- function(values) {
  raster <- template
  terra::values(raster) <- values
  raster
}

# Four irregular parks embedded in an urban matrix, plus a simple road grid.
parks <- pmin(1, exp(-((x - 0.20)^2 + (y - 0.76)^2) / 0.012) +
                 exp(-((x - 0.72)^2 + (y - 0.72)^2) / 0.020) +
                 exp(-((x - 0.30)^2 + (y - 0.25)^2) / 0.010) +
                 exp(-((x - 0.78)^2 + (y - 0.30)^2) / 0.014))
roads <- as.numeric(abs(sin(8 * pi * x)) < 0.08 |
                    abs(sin(8 * pi * y)) < 0.08)
canopy <- pmin(1, pmax(0, 0.08 + 0.90 * parks - 0.20 * roads +
                          rnorm(terra::ncell(template), sd = 0.04)))
impervious <- pmin(1, pmax(0, 0.15 + 0.25 * (1 - parks) + 0.40 * roads + 0.25 * x +
                              rnorm(terra::ncell(template), sd = 0.04)))
basis_stack <- create_basis_stack(list(
  canopy = layer(canopy),
  impervious = layer(impervious)
), rescale = TRUE)

# Smooth white noise twice to create a spatially autocorrelated log-intensity
# effect. It is deliberately omitted from the fitted model.
spatial_effect <- terra::focal(layer(rnorm(terra::ncell(template))),
                               w = matrix(1, 9, 9), fun = "mean",
                               na.rm = TRUE, fillvalue = 0)
spatial_effect <- terra::focal(spatial_effect, w = matrix(1, 9, 9),
                               fun = "mean", na.rm = TRUE, fillvalue = 0)
spatial_effect <- 0.35 * as.numeric(scale(terra::values(spatial_effect)))

true_log_resistance <- -1 * canopy + 1 * impervious
true_resistance <- layer(exp(true_log_resistance))
ds_torch_setup()
circuit_solver <- DiffiScape:::.ds_env$torch_module$`_get_circuit_module`()
truth <- circuit_solver$solve_circuit_absorption(
  terra::as.matrix(true_resistance, wide = TRUE), absorption = 0.20,
  source_spacing = 1L, source_from_resistance = TRUE
)
true_connectivity <- layer(as.numeric(reticulate::py_to_r(truth$current_density)))
log_lambda <- 2 * log1p(terra::values(true_connectivity)) + spatial_effect
sampling_weight <- exp(log_lambda - max(log_lambda))
obs_cells <- sample(seq_len(terra::ncell(template)), 1000L,
                    replace = TRUE, prob = sampling_weight)
obs_points <- coords[obs_cells, c("x", "y")]
output_dir <- file.path(tempdir(), "diffiscape-torch-mlp")
```

## Inspect the simulated truth

```{r simulated-truth, fig.width=12, fig.height=4, dpi=96}
op <- par(mfrow = c(1, 3))
plot(basis_stack[["canopy"]], main = "Canopy cover")
plot(basis_stack[["impervious"]], main = "Impervious surface")
plot(true_connectivity, main = "True connectivity")
par(op)
```

## Verify gradients and fit a compact MLP

The gradient check uses a 12 x 12 centre crop. The model itself runs on the
full 100 x 100 landscape, but uses a narrow one-hidden-layer MLP and at most
10 epochs with a relatively local absorption term to remain practical on a
weak CPU. The short run uses a conservative sign-correct warm start
`c(0, -0.5, 0.5)`, rather than the true coefficients, so the example does not
mistake a cold-start numerical transient for an ecological conclusion.

```{r fit}
gradient_check <- verify_torch_gradient(
  basis_stack, crop_size = 12L, solver = "global_absorption", absorption = 0.20
)
stopifnot(isTRUE(gradient_check$pass))

cfg <- default_torch_config()
cfg$solver <- "global_absorption"
cfg$absorption <- 0.20
cfg$source_spacing <- 1L
cfg$model_type <- "mlp"
cfg$hidden_dim <- 4L
cfg$n_hidden_layers <- 1L
cfg$lr <- 0.005
cfg$n_epochs <- 10L
cfg$patience <- 5L
cfg$warmup_epochs <- 1L
cfg$reg_mean <- 0.2
cfg$reg_var <- 0.0
cfg$reg_skip <- 0.05
cfg$log_R_baseline <- 0.0
cfg$seed <- 42L
cfg$device <- "cpu"

results <- run_torch_pipeline(
  basis_stack = basis_stack, obs_points = obs_points,
  config = cfg, output_dir = output_dir,
  warm_start_theta = c(0, -0.5, 0.5)
)
```

## Diagnostic and visual checks

The fitted model intentionally omits the simulated spatial random effect, so
it should recover broad resistance/connectivity structure rather than every
local intensity fluctuation. The assertions catch numerical failures; the
loss trace and side-by-side maps make broad agreement visually inspectable.

```{r diagnostics, fig.width=12, fig.height=4, dpi=96}
stopifnot(
  all(is.finite(results$loss_history)),
  all(is.finite(terra::values(results$resistance_raster))),
  all(is.finite(terra::values(results$connectivity_raster))),
  all(is.finite(terra::values(results$intensity_raster))),
  all(terra::values(results$resistance_raster) > 0),
  all(terra::values(results$intensity_raster) > 0)
)
print(data.frame(
  coefficient = c("canopy", "impervious"),
  true_log_resistance_effect = c(-1, 1),
  fitted_effective_loglinear = results$effective_loglinear[-1]
))

plot(results$loss_history, type = "l", xlab = "Epoch", ylab = "Objective",
     main = "Training objective")
op <- par(mfrow = c(1, 3))
plot(true_resistance, main = "True resistance")
plot(results$resistance_raster, main = "Fitted resistance")
plot(results$connectivity_raster, main = "Fitted connectivity")
par(op)
```

```{r save}
saveRDS(
  list(
    true_log_resistance = true_log_resistance,
    true_connectivity = true_connectivity,
    effective_loglinear = results$effective_loglinear,
    best_loglik = results$best_loglik,
    best_epoch = results$best_epoch,
    total_time = results$total_time
  ),
  file.path(output_dir, "torch_optimization_results.rds")
)
```
