---
title: "Neural Network Resistance via PyTorch (MLP / ConvNet)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Neural Network Resistance via PyTorch (MLP / ConvNet)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

This vignette demonstrates the PyTorch neural-network pipeline. Instead of a
log-linear resistance parameterization, a learned MLP (or dilated ConvResNet)
maps the environmental basis functions to log-resistance at every pixel,
capturing nonlinear covariate effects and spatial interactions. The global
absorption circuit solver runs the connectivity forward pass; PyTorch autograd
differentiates back through it.

This is the fastest pipeline per iteration (~50–100x vs. GAM), and it can fit
models with hundreds to thousands of parameters.

This pipeline requires a 5-layer basis stack: the four NLCD layers plus an
elevation layer from the Copernicus DEM (or any DEM).

## Setup

```{r paths}
library(DiffiScape)
library(terra)

obs_data_path  <- "path/to/train_points.rds"
nlcd_dir       <- "path/to/nlcd_data/"
elevation_path <- "path/to/elevation.tif"   # Copernicus DEM or similar
output_dir     <- "output/torch"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load data and build 5-layer basis

Elevation is included as a fifth covariate. `ds_create_basis()` accepts a
character vector of paths to individual rasters and aligns them to a common
grid automatically.

```{r load-data}
obs_points <- readRDS(obs_data_path)

nlcd_paths <- list.files(nlcd_dir, pattern = "\\.tif$", full.names = TRUE)
basis_stack <- ds_create_basis(c(nlcd_paths, elevation_path), rescale = TRUE)
cat("Basis layers:", names(basis_stack), "\n")   # should be 5 layers
```

## Initialise PyTorch

```{r init-torch}
ds_torch_setup()   # initialises reticulate + Python environment
ds_torch_check()   # confirms torch is available
```

## Verify gradients (optional but recommended)

End-to-end finite-difference check of the circuit–NN gradient chain. A
relative error above ~5 % suggests a numerical issue worth investigating
before committing to a full training run.

```{r verify-gradients}
verify_torch_gradient(basis_stack,
                      solver     = "global_absorption",
                      absorption = 0.002)

verify_conv_gradient(basis_stack,
                     solver        = "global_absorption",
                     absorption    = 0.002,
                     conv_channels = 16L,
                     n_conv_layers = 3L)
```

## Configure the torch pipeline

Settings match the original 06d analysis. Key choices:

- `solver = "global_absorption"` — one global circuit solve with an absorption
  term; much faster than 23 000 local Omniscape solves.
- `use_conv = TRUE` — dilated ConvResNet captures spatial context.
- `intensity_hidden = 16L` — learned intensity MLP instead of parametric
  γ·log(1+C).

```{r config}
cfg <- default_torch_config()

# Solver
cfg$solver           <- "global_absorption"
cfg$absorption       <- 0.002
cfg$focal_fraction   <- 0.5
cfg$source_spacing   <- 1L
cfg$source_from_resistance <- TRUE

# ConvResNet architecture
cfg$hidden_dim       <- 16L
cfg$n_hidden_layers  <- 3L
cfg$use_conv         <- TRUE
cfg$n_conv_layers    <- 3L
cfg$conv_channels    <- 16L
cfg$conv_kernel_size <- 3L
cfg$dropout          <- 0.1
cfg$use_dilated      <- TRUE
cfg$intensity_hidden <- 16L   # 0 = parametric; >0 = learned MLP

# Optimiser
cfg$lr               <- 0.05
cfg$weight_decay     <- 1e-4
cfg$n_epochs         <- 1000L
cfg$patience         <- 100L
cfg$grad_clip        <- 10.0
cfg$warmup_epochs    <- 10L

# Regularisation
cfg$reg_mean         <- 1.0    # penalises deviation from log_R_baseline
cfg$reg_var          <- 0.0
cfg$reg_skip         <- 0.1    # skip-connection penalty (prevents NN collapse)
cfg$log_R_baseline   <- 3.0    # target: exp(3) ≈ 20

cfg$seed             <- 42L
cfg$device           <- "auto"  # uses GPU if available
```

## Run the torch pipeline

```{r run-torch}
results <- run_torch_pipeline(
  basis_stack = basis_stack,
  obs_points  = obs_points,
  config      = cfg,
  output_dir  = output_dir
)

cat("Best log-likelihood:", results$best_loglik, "\n")
cat("Best epoch:", results$best_epoch, "/", results$n_epochs_run, "\n")
cat("Total time:", results$total_time, "sec\n")
cat("Effective log-linear params:\n")
print(results$effective_loglinear)   # [r_0, z_1, ..., z_5]
```

## Inspect outputs

```{r outputs}
terra::writeRaster(results$resistance_raster,
                   file.path(output_dir, "optimized_resistance.tif"), overwrite = TRUE)
terra::writeRaster(results$connectivity_raster,
                   file.path(output_dir, "optimized_connectivity.tif"), overwrite = TRUE)
terra::writeRaster(results$intensity_raster,
                   file.path(output_dir, "predicted_intensity.tif"), overwrite = TRUE)
```

## Optional: post-hoc GAM on the optimised connectivity surface

Fitting a spatial GAM after training allows you to inspect smooth covariate
effects and spatial residuals without re-running the circuit solver.

```{r posthoc-gam}
icfg <- default_intensity_config()
icfg$covariate_type     <- "smooth"
icfg$include_spatial_re <- TRUE

intensity_fit <- ds_fit_intensity(
  opt_result         = results,
  basis_stack        = basis_stack,
  obs_points         = obs_points,
  omniscape_settings = list(radius = 15L, block_size = 5L),
  intensity_config   = icfg
)

diag <- ds_diagnose(intensity_fit, obs_points, results$connectivity_raster,
                    intensity_config = icfg, plot = TRUE)
```

## Save summary

```{r save}
saveRDS(
  list(
    effective_loglinear = results$effective_loglinear,
    alpha               = results$alpha,
    gamma               = results$gamma,
    best_loglik         = results$best_loglik,
    best_epoch          = results$best_epoch,
    n_params            = results$n_params,
    total_time          = results$total_time
  ),
  file.path(output_dir, "torch_optimization_results.rds")
)
```
